---
title: "Women in Parliament - Tidy Data"
author: "Filipe Santos"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
```

```{r imageStart, echo=FALSE, out.width="20%"}
knitr::include_graphics("images/Women_in_Parliament_hex.svg")
```


# Objectives

*Explore the geographical and time trends for the percentage of women in 
national parliaments.*

# Understanding the Data

## The World Bank Data 

The raw data for *"Proportion of seats held by women in national parliaments"* 
includes the percentage of women in parliament (_"single or lower parliamentary chambers 
only"_) by country (region) and year. It can be downloaded from:

- https://data.worldbank.org/indicator/SG.GEN.PARL.ZS 

As part of its "open data" mission the World Bank offers _"free and open access 
to global development data"_ kindly licensed under the "Creative Commons Attribution 
4.0 (CC-BY 4.0)".

## Source Data

The data originates from the "Inter-Parliamentary Union" (IPU)[^ipuwebsite] which
provides an *_"Archive of statistical data on the percentage of women in 
national parliaments"_* going back to 1997 on a monthly basis:

- http://archive.ipu.org/wmn-e/classif-arc.htm

The World Bank data is for “single or lower parliamentary chambers only”, while 
the IPU also presents data for “Upper Houses or Senates”. Moreover, the IPU provides 
the actual numbers used to calculate the percentages (which the World Bank does not).

## Data limitations

Take caution when interpreting these data, as parliamentary systems vary from country 
to country, and in some cases over time. Some of the issues to consider include:

- Who has, and who does not have, the right to become a Member of Parliament (MP)?
- How does someone become an MP? Through democratic elections? How is "democratic
election" defined? 
- What is the real power of MPs and their parliament? Can MPs make a difference?

## Data definitions & assumptions

### "Women"

The definition for “women” is not given, so we will assume that it refers to a 
binary classification for gender (sex).

### "Country (Region)"

The definition of countries and regions can change over time.  (e.g. formation of new
countries after conflicts, new member states joining a pre-existing collective). 
How are these changes reflected in the data? How do they affect the interpretation? 

## Pro tip

Understand the limitations of your data before anybody else points them out to you. 

# About the data file

The data is stored in a file called: ``r list.files(path = here("data"), pattern = "^API")``

To simplify things we have copied it to `WB-WiP.csv` (which also allows us to 
maintain the original file in case something goes wrong).

## Pro tip

Always keep a backup copy of the data. Alternatively, set the data file(s) to 
“read-only” to protect it from being overwritten or modified.

# Importing the data

Based on our findings above, we can "skip" the first four lines and treat the
fifth line as column (variable) names. Also note the use of the `check.names`
argument to ensure that the column names are compliant in R.

```{r readData, collapse=TRUE}
library(data.table)
library(here)
wip <- fread(here("data", "WB-WiP.csv"), 
             skip = 4, header = TRUE,
             check.names = TRUE)
wip
```

# Data Wrangling Aims

## Superflous Columns

```{r NAcheck}
wip[, .N, by=.(V65)]
wip[, .N, by=.(Indicator.Name)]
wip[, .N, by=.(Indicator.Code)]

```

## Removing Columns

```{r RemoveC}
wip[, c("Indicator.Name", "Indicator.Code", "V65"):=NULL]
setnames(wip, c("Country.Name", "Country.Code"), c("Country", "Code"))
head(names(wip))
tail(names(wip))
```


## Reshape to long format.

```{r Longformat}
WP <- melt(wip,
           id.vars = c("Country", "Code"),
           measure = patterns("^X"),
           variable.name = "YearC",
           value.name = c("pctWiP"),
           na.rm = TRUE)
WP
```

## Final tweaks to WP

```{r yearratio}
WP[, `:=`(Year=as.numeric(gsub("[^[:digit:].]",
                               "", YearC)),
          Ratio = (100-pctWiP)/pctWiP)][
            , YearC := NULL]
setcolorder(WP, c("Country", "Code", "Year", "pctWiP", "Ratio"))

WP
```

# Exploratory Analysis

## Time trends for South Africa

```{r SAfrdata}
WP[Country %in% "South Africa"]
```

## Visualization

```{r SAfrWiP}
library(ggplot2)
library(magrittr)
WP[Country %in% "South Africa"] %>%
ggplot(aes(Year, pctWiP)) + 
  geom_line() + geom_point() + 
  scale_y_continuous(limits=c(0,50)) +
  ylab("% Women in Parliament")
```
### Interpretation

Huge jump in number of Women in Parliament from 1990 to 1997. Steady levels through the 2000's, similar to Portugal's current levels. 2009 increased WiP percentage to the 40's, with a steady % to today, higher than Portugal's current numbers.

```{r imageEnd, echo=FALSE, out.width="60%"}
knitr::include_graphics("images/Women_in_Parliament_rect.svg")
```

## Portugal vs EU countries

```{r PTvsEU}
WP[Country %in% c("Portugal", "Sweden", "Spain", "Hungary", "Romania", "Finland", "Germany", "European Union")] %>%
  ggplot(aes(Year, pctWiP, colour=Country)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks=seq(1990,2020,5)) +
  scale_y_continuous(limits = c(0,50),
                     breaks=seq(0,50,by=10)) +
  ggtitle("Women in Parliament: EU Countries") +
  ylab("% Women in Parliament")
```

##South Africa vs other countries

```{r sAfrVs}
WP[Country %in% c("South Africa", "Sweden", "Portugal", "Venezuela", "Russia", "China", "Japan", "European Union")] %>%
  ggplot(aes(Year, pctWiP, colour=Country)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks=seq(1990,2020,5)) +
  scale_y_continuous(limits = c(0,50),
                     breaks=seq(0,50,by=10)) +
  ggtitle("Women in Parliament: EU Countries") +
  ylab("% Women in Parliament")
```

##Highest percentage of women

```{r mostWiP}
WP[order(-pctWiP), head(.SD, 10)]
```

##Highest WiP per year

```{r wipPerYear}
WP[order(Year, -pctWiP), head(.SD, 1), by = Year]
```

##Adding continent names

```{r addCont}
library(countrycode)
c1 <- as.data.table(codelist)[, .(continent, wb)]
setnames(c1, c("continent", "wb"),
         c("Continent", "Code"))
cWP <- c1[WP, on="Code"]
cWP
```

##Highest %WiP by year and continent

```{r %WiPYC}
cWP[Year %in% c(1990, 2018) & !is.na(Continent)][
  order(Year, -pctWiP), head(.SD, 1),
  by = .(Year, Continent)][
    order(Continent, Year),
    .(Continent, Year, Country, pctWiP)]
```

##Decline in Percentage

```{r decl%}
dWP <- cWP[
  order(Country, Year), .SD[c(1, .N)],
  by=Country][,
              pctDiff := pctWiP - shift(pctWiP), by=Country][
                pctDiff<0][
                  order(pctDiff)]
dWP[!is.na(Continent),
    .(Country, pctWiP, pctDiff)]
```

##5+% Decline Visualization

```{r 5%Decl}


dclpct <- unique(dWP[!is.na(Continent) &
                       pctDiff <= -5]$Country)

WP[Country %in% dclpct] %>%
  ggplot(aes(Year, pctWiP, colour=Country)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks=seq(1990, 2020, 5)) +
  scale_y_continuous(limits=c(0,40),
                     breaks=seq(0, 40, by=10)) +
  ggtitle("Women in Parliament: Decline >=5%") +
  ylab("% Women in Parliament")
```

##Global ranks by Year

```{r gRankYear}
cWP[!is.na(Continent),
    `:=`(RankG = rank(-pctWiP), TotalG = .N),
    by = .(Year)]
```

##Global ranking - Portugal

```{r ptRank}
cWP[Country == "Portugal",
    .(Country, Year, pctWiP, Ratio, RankG, TotalG)][
      order(Year)]
```

##Global ranking - South Africa

```{r sAfrRank}
cWP[Country == "South Africa",
    .(Country, Year, pctWiP, Ratio, RankG, TotalG)][
      order(Year)]
```
Values for 1990 are low but since 1997 South Africa has consistently been one of the top countries in regards to percentage of women in Parliament, having never dropped from the top 15. From 2009 (where it ranked the highest, 3rd) to 2018, South Africa has always been in the top 10, putting aside 2014 where it ranked 11th. 
Except for 1990, South Africa has always ranked higher than Portugal.

##Continent ranks by year

```{r contRankYear}
cWP[!is.na(Continent),
    `:=`(RankC = rank(-pctWiP), TotalC = .N),
    by = .(Continent, Year)]
```

##Portugal's ranking in Europe

```{r ptRankEU}
cWP[Country=="Portugal",
    .(Country, Year, pctWiP, Ratio, RankC, TotalC)][order(Year)]
```

##Plot of Portugal's ranking vs 6 other EU countries

```{r PTvsEUplot}
cWP[Country %in% c("Portugal", "Sweden", "Spain", "Hungary", "Romania", "Finland", "Germany")] %>%
  ggplot(aes(Year, RankC, colour=Country)) +
  geom_line() + 
  geom_point() + 
  scale_x_continuous(breaks=seq(1990, 2020, 5)) + 
  scale_y_continuous(limits=c(0,45),
                     breaks=seq(0, 45, by=10)) +
  ggtitle("Women in Parliament: Ranked") + 
  ylab("Rank in Europe")
```

##South Africa ranking in Africa

```{r sAfrRank}
cWP[Country=="South Africa",
    .(Country, Year, pctWiP, Ratio, RankC, TotalC)][order(Year)]
```

##South Africa ranking vs other African countries

```{r SAfrvsAfrplot}
cWP[Country %in% c("South Africa", "Rwanda", "Guinea-Bissau", "Angola", "Mozambique", "Morocco", "Tunisia")] %>%
  ggplot(aes(Year, RankC, colour=Country)) +
  geom_line() + 
  geom_point() + 
  scale_x_continuous(breaks=seq(1990, 2020, 5)) + 
  scale_y_continuous(limits=c(0,45),
                     breaks=seq(0, 45, by=10)) +
  ggtitle("Women in Parliament: Ranked") + 
  ylab("Rank in Africa")
```

South Africa has always been in the top four countries (with a number 1 position in 98 and 99), if we put aside its values for 1990.

##Highest rank by year and continent

```{r highRankYC}
cWP[Year %in% c(1990,2018) & RankC==1][
  order(Continent, Year),
  .(Continent, Year, Country, pctWiP, RankC)]
```

##Overall picture - Global trends

```{r globalTrends}
library(gghighlight)
cWP[is.na(Continent)] %>%
  ggplot(aes(Year, pctWiP, group=Country)) +
  geom_line() +
  gghighlight(Country=="World",
              use_direct_label = FALSE) + 
  scale_x_continuous(breaks=seq(1990, 2020, 5)) +
  scale_y_continuous(limits=c(0,40),
                     breaks=seq(0,40,by=10)) +
  ggtitle("Women in Parliament: Global Trends") +
  ylab("% Women in Parliament")
```

```{r sessInfo}
sessionInfo()
```

